\Large\bf Scalable and Probabilistic Leaderless BFT Consensus through Metastability

### 2020-11-05 paper_fr.tex
French translation of section about Slush protocol

\subsection{Slush: Introducing Metastability}
The core of our approach is a single-decree consensus protocol,
inspired by epidemic or gossip protocols.
% In this approach, a state in which all possible outcomes are equally
% probable is metastable and disturbances are likely to cause the
% protocol to converge to one of multiple stable states.
% We will call such consensus protocols metastable themselves.
%We can model a preference for a decision as a kind of infection: a node catches a disease, and converts other nodes to carry the same disease in epidemic fashion.
%If a threshold of its interactions are with other infected hosts, it flips its color to the competing agent and joins their ranks.
%
The simplest protocol, Slush, is the foundation of this family, shown in Figure~\ref{fig:slush-loop}.
Slush is \emph{not} tolerant to Byzantine faults, only crash-faults (CFT), but serves as an illustration for the BFT protocols that follow.
For ease of exposition, we will describe the operation of Slush using a decision between two conflicting colors, red and blue.
% , though the technique generalizes to any number of competing proposed values.
% Slush operates as follows.

In Slush, a node starts out initially in an uncolored state.
Upon receiving a transaction from a client, an uncolored node updates its own color to the one carried in the transaction and initiates a query.
To perform a query, a node picks a small, constant sized ($k$) sample of the network uniformly at random, and sends a query message.
Upon receiving a query, an uncolored node adopts the color in the query, responds with that color, and initiates its own query, whereas a colored node simply responds with its current color.
% If $k$ responses are not received within a time bound, the node picks an additional sample from the remaining nodes uniformly at random and queries them until it collects all responses.
Once the querying node collects $k$ responses, it checks if a fraction $\ge\alpha$ are for the same color, where $\alpha > \floor{k/2}$ is a protocol parameter.
If the $\alpha$ threshold is met and the sampled color differs from the node's own color, the node flips to that color.
It then goes back to the query step, and initiates a subsequent round of query, for a total of $m$ rounds.
Finally, the node decides the color it ended up with at time $m$.

@@

\newcommand{\codecolor}{\mathit{col}}
\newcommand{\codecount}{\mathit{cnt}}
\newcommand{\codelastcol}{\mathit{lastcol}}
\newcommand{\assign}{\coloneqq}
\begin{figure}
    \small
\begin{algorithmic}[1]
    \Procedure{onQuery}{$v, \codecolor'$}
        \IIf{$\codecolor = \bot$} $\codecolor \assign \codecolor'$
        \State\Call{respond}{$v, \codecolor$}
    \EndProcedure
    \Procedure{slushLoop}{$u, \codecolor_0 \in \{\texttt{R}, \texttt{B}, \bot\}$}
        \State $\codecolor \assign \codecolor_0$ \textrm{// initialize with a color}
        \For{$r \in \{1\ldots m\}$}
            \State \textrm{// if $\bot$, skip until \textsc{onQuery} sets the color}
            \If{$\codecolor = \bot$}
                \Continue
            \EndIf
            \State \textrm{// randomly sample from the known nodes}
            \State $\mathcal{K} \assign \Call{sample}{\mathcal{N}\backslash u, k}$
            \State $P \assign \texttt{[}\Call{query}{v, \codecolor}\quad\textbf{for}\ v \in \mathcal{K}\texttt{]}$
            \For{$\codecolor' \in \{\texttt{R}, \texttt{B}\}$}
                \If{$P.\Call{count}{\codecolor'} \ge \alpha$}
                    \State $\codecolor \assign \codecolor'$
                \EndIf
            \EndFor
        \EndFor
        \State \Call{accept}{$\codecolor$}
    \EndProcedure
    \captionof{figure}{Slush protocol. Timeouts elided for readability.}\label{fig:slush-loop}
\end{algorithmic}
\end{figure}

Slush has a few properties of interest. 
First, it is almost \emph{memoryless}: a node retains no state between rounds other than its current color, and in particular maintains no history of interactions with other peers.
Second, unlike traditional consensus protocols that query every participant, every round involves sampling just a small, constant-sized slice of the network at random.
Third, Slush makes progress under any network configuration (even fully bivalent state, i.e. 50/50 split between colors), since random perturbations in sampling will cause one color to gain a slight edge and repeated samplings afterwards will build upon and amplify that imbalance.
Finally, if $m$ is chosen high enough, Slush ensures that all nodes will be colored identically with high probability (whp).
Each node has a constant, predictable communication overhead per round, and $m$ grows logarithmically with $n$.

\tronly{
The Slush protocol does not provide a strong safety guarantee in the presence of Byzantine nodes.
In particular, if the correct nodes develop a preference for one color, a Byzantine adversary can attempt to flip nodes
to the opposite so as to keep the network in balance, preventing a decision.
We address this in our first BFT protocol that introduces more state storage at the nodes.
}{
We next examine how to extend Slush to tolerate Byzantine behavior.
}

>>>>

%0 \subsection{Slush: Introducing Metastability}
\subsection{Slush: A propos de métastabililté}

%O The core of our approach is a single-decree consensus protocol,
%O inspired by epidemic or gossip protocols.
%O The simplest protocol, Slush, is the foundation of this family, shown in Figure~\ref{fig:slush-loop}.
%O Slush is \emph{not} tolerant to Byzantine faults, only crash-faults (CFT), but serves as an illustration for the BFT protocols that follow.
%O For ease of exposition, we will describe the operation of Slush using a decision between two conflicting colors, red and blue.

Le principe de notre approche est un protocole de consensus à décision simple inspiré des protocoles dits "épidémiques"
ou "gossip". Le protocole le plus simple, Slush, est la fondation de cette famille, comme le montre la
Figure~\ref{fig:slush-loop}. Slush \emph{n'est pas} tolérant aux fautes byzantines, seulement aux fautes par défaut
de fonctionnement (CFT), mais sert à mettre en avant les autres protocoles tolérants aux fautes byzantines qui vont
suivre. Dans un souci de simplicité de présentation, nous décrirons le fonctionnement de Slush en le comparant
à une décision à prendre entre deux couleurs incompatibles, le rouge et le bleu.

%O In Slush, a node starts out initially in an uncolored state.
%O Upon receiving a transaction from a client, an uncolored node updates its own color to the one carried in the transaction and initiates a query.
%O To perform a query, a node picks a small, constant sized ($k$) sample of the network uniformly at random, and sends a query message.
%O Upon receiving a query, an uncolored node adopts the color in the query, responds with that color, and initiates its own query, whereas a colored node simply responds with its current color.
%O Once the querying node collects $k$ responses, it checks if a fraction $\ge\alpha$ are for the same color, where $\alpha > \floor{k/2}$ is a protocol parameter.
%O If the $\alpha$ threshold is met and the sampled color differs from the node's own color, the node flips to that color.
%O It then goes back to the query step, and initiates a subsequent round of query, for a total of $m$ rounds.
%O Finally, the node decides the color it ended up with at time $m$.

Avec Slush, un noeud démarre dans un état de couleur neutre. Après réception d'une transaction envoyée par un client,
un node de couleur neutre adopte la couleur définie dans la transaction et initie une requête. Pour lancer une
requête, le noeud choisit un échantillon réduit et de taille constante ($k$) de noeuds appartenant au réseau suivant
une distribution uniformément aléatoire puis initie sa propre requête, tandis qu'un noeud déja coloré répond
simplement avec sa couleur actuelle. Une fois que le noeud qui fait la requête récupère $k$ réponses, il vérifie si
une proportion $\ge\alpha$ de l'échantillon retourne la même couleur, $\alpha > \floor{k/2}$ étant un paramètre
du protocole. Si le seuil $\alpha$ est dépassé et que la couleur dominante revenant de l'échantillon sollicité est
différente de la couleur propre du node à l'origine de la requête, ce node adopte cette nouvelle couleur. Il revient
ensuite à l'étape de génération de la requête et lance un nouveau tour de requête jusqu'à un total de $m$ tours.
En définitive, le node décide de sa couleur finale à l'instant $m$.

@@

\newcommand{\codecolor}{\mathit{col}}
\newcommand{\codecount}{\mathit{cnt}}
\newcommand{\codelastcol}{\mathit{lastcol}}
\newcommand{\assign}{\coloneqq}
\begin{figure}
    \small
\begin{algorithmic}[1]
    \Procedure{onQuery}{$v, \codecolor'$}
        \IIf{$\codecolor = \bot$} $\codecolor \assign \codecolor'$
        \State\Call{répondre}{$v, \codecolor$}
    \EndProcedure
    \Procedure{slushLoop}{$u, \codecolor_0 \in \{\texttt{R}, \texttt{B}, \bot\}$}
        \State $\codecolor \assign \codecolor_0$ \textrm{// initialise avec une couleur}
        \For{$r \in \{1\ldots m\}$}
            \State \textrm{// Si $\bot$, réitère jusqu'à ce que \textsc{onQuery} applique la couleur}
            \If{$\codecolor = \bot$}
                \Continue
            \EndIf
            \State \textrm{// échantillonne aléatoirement parmi les noeuds connus}
            \State $\mathcal{K} \assign \Call{échantillon}{\mathcal{N}\backslash u, k}$
            \State $P \assign \texttt{[}\Call{requête}{v, \codecolor}\quad\textbf{pour}\ v \in \mathcal{K}\texttt{]}$
            \For{$\codecolor' \in \{\texttt{R}, \texttt{B}\}$}
                \If{$P.\Call{nombre}{\codecolor'} \ge \alpha$}
                    \State $\codecolor \assign \codecolor'$
                \EndIf
            \EndFor
        \EndFor
        \State \Call{accepte}{$\codecolor$}
    \EndProcedure
    \captionof{figure}{Protocole Slush. Les temps d'expiration sont omis dans un souci de lisibilité.}\label{fig:slush-loop}
\end{algorithmic}
\end{figure}

%O Slush has a few properties of interest.
%O First, it is almost \emph{memoryless}: a node retains no state between rounds other than its current color, and in particular maintains no history of interactions with other peers.
%O Second, unlike traditional consensus protocols that query every participant, every round involves sampling just a small, constant-sized slice of the network at random.
%O Third, Slush makes progress under any network configuration (even fully bivalent state, i.e. 50/50 split between colors), since random perturbations in sampling will cause one color to gain a slight edge and repeated samplings afterwards will build upon and amplify that imbalance.
%O Finally, if $m$ is chosen high enough, Slush ensures that all nodes will be colored identically with high probability (whp).
%O Each node has a constant, predictable communication overhead per round, and $m$ grows logarithmically with $n$.

Slush a quelques propriétés intéréssantes. Premièrement, il n'a quasiment \emph{pas besoin de mémoire}: un noeud ne
stocke aucun état intermédiaire entre chaque tour autre que sa couleur actuelle, et plus particulièrement ne stocke
aucun historique de ses interactions avec les autres noeuds. Deuxièmement, contrairement aux protocoles de consensus
traditionnels qui sollicitent tous les participants, chaque tour n'implique qu'un échantillon du réseau réduit et de
taille constante choisi aléatoirement. Troisièmement, Slush progresse quelque soit la configuration du réseau (même
depuis un état entièrement bivalent, ex: partage 50/50 entre les deux couleurs) vu que des perturbations aléatoires
sur l'échantillonnage auront pour conséquence de fournir un léger avantage à l'une des deux couleurs qui sera ensuite
amplifié au fur et à mesure des échantillonnages.
Finalement si $m$ est choisi pour être suffisamment élevé, Slush s'assure que tous les noeuds convergeront vers la
même couleur avec une haute probabilité. Chaque noeud nécéssite un temps de communication par tour constant et
prévisible, et $m$ croît de manière logarithmique suivant $n$.

\tronly{
%O The Slush protocol does not provide a strong safety guarantee in the presence of Byzantine nodes.
%O In particular, if the correct nodes develop a preference for one color, a Byzantine adversary can attempt to flip nodes
%O to the opposite so as to keep the network in balance, preventing a decision.
%O We address this in our first BFT protocol that introduces more state storage at the nodes.

Le protocole Slush ne fournit pas de fortes garanties quant à la sûreté de fonctionnement en la présence de noeuds
byzantins. En particulier si les noeuds légitimes développent une préférence pour une certaine couleur, un adversaire
byzantin pourra inverser la couleur des noeuds afin de maintenir le système à l'équilibre, incapable de prendre une
décision. Nous adressons ce problème dans notre premier protocole tolérant aux fautes byzantines qui introduit un
meilleur suivi de l'état au niveau des nodes.
}{
%O We next examine how to extend Slush to tolerate Byzantine behavior.
Nous examinons ensuite comment améliorer Slush pour être tolérant au problème des généraux byzantins.
}
